A Review of Mortality Models

### 2020-09-13 mortality-models/draft.tex
WIP 13Sep20

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\numberwithin{equation}{section}
\DeclareMathOperator{\pr}{P}
\linespread{1.3}
\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behavior of a member of a large
population one can often make statistical statements about the same population 
with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one can reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they to describe how deaths occur in a given society are
called `mortality models'.

@@

\section{Stochastic cohort models}\label{s9}
The stochastic models of Lee-Carter and Cairns-Blake-Dowd were unable to
incorporate cohort effects. To remedy this shortcoming, Renshaw and Haberman
\cite{renshaw2006cohort} proposed a modification of the Lee-Carter model that 
includes cohorts. It can be expressed mathematically as
\begin{equation}\label{s9e1}
\log{M}(x, t) = {a}_x + {b}_x{k}_t + {c}_x\gamma_{t-x},
\end{equation}
where the symbols have the same meaning as in Lee-Carter model. In the new
term, third on the right hand side, $\gamma_{t-x}$ captures the cohort effect.
It is a function of $t - x$, the (approximate) year of birth and ${c}_x$
is just the regression coefficient of this term. Although the model fitted the
male mortality data in England and Wales quite well it turned to not as 
robust as other models. The lack of robustness is believed to be related to
the shape of the likelihood function. It has more than one local maximum.
Therefore, when we change the age or the year range the optimizer ends up
settling on a different maximum each time\cite{cairns2008modelling}. Currie
simplified the Renshaw-Haberman model to
\begin{equation}\label{s9e2}
\log{M}(x, t) = {a}_x + {k}_t + \gamma_{t-x}.
\end{equation}
The simplification solved the problem of robustness but made the fit poorer
\cite{plat2009stochastic}. Plat introduced a model to address the deficiencies
of all the previous models while retaining their benefits 
\cite{plat2009stochastic}. Its mathematical structure is
\begin{equation}\label{s9e3}
\log M(x,t) = a_x + \kappa^1_t + \kappa^2_t(\bar{x} - x) + 
\kappa^3_t(\bar{x} - x)^+ + \gamma_{t-x},
\end{equation}
where $(\bar{x} - x)^+ = \max(\bar{x} - x, 0)$ and $\bar{x}$ is the mean age
in the sample range. The term $a_x$ plays a 
similar role as it did in Lee-Carter model. It controls the general shape
of the mortality curve. Of the four stochastic factors, $\kappa_t^1$
represents the change in the level of mortality across all ages, $\kappa_t^2$
represents the change in the level of mortality between ages, $\kappa_t^3$
captures the differences in mortality behavior in the early adulthood stage
and $\gamma_{t-x}$ captures the cohort effect. If one is interested in 
modeling only the advanced ages then one need not worry about the `accident
hump' of young adulthood and use the simplified model,
\begin{equation}\label{s9e4}
\log M(x,t) = a_x + \kappa^1_t + \kappa^2_t(\bar{x} - x) + \gamma_{t-x},
\end{equation}
Like the Lee-Carter model, this one too suffers from the `identifiability
problem'. For constants $c_1, c_2, d$, the set of parameters
\begin{eqnarray*}
\bar{\gamma}_{t-x} &=& \gamma_{t-x} + c_1 + c_2(t - x) \\
\bar{\kappa}_t^1 &=& \kappa^1_t + c_1 - d\bar{x}c_2 \\
\bar{\kappa}_t^2 &=& \kappa^2_t + dc_2 \\
\bar{a}_x &=& a_x + (1 - d)c_2
\end{eqnarray*}
also lead the the same values for $\log M(x, t)$. The problem is mitigated 
by requiring that 
\begin{eqnarray*}
\sum_{i=x_0}^{x_1} \gamma_i &=& 0 \\
\sum_{i=x_0}^{x_1} i\gamma_i &=& 0 \\
\sum_t\kappa_t^3 &=& 0,
\end{eqnarray*}
where $x_0$ and $x_1$ are the earliest and the latest years of birth to which
the cohort effect is fitted.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\usepackage{url}
\numberwithin{equation}{section}
\DeclareMathOperator{\pr}{P}
\DeclareMathOperator{\E}{E}
\linespread{1.3}
\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behavior of a member of a large
population one can often make statistical statements about the same population 
with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one can reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they to describe how deaths occur in a given society are
called `mortality models'.

@@

\section{Stochastic cohort models}\label{s9}
The stochastic models of Lee-Carter and Cairns-Blake-Dowd were unable to
incorporate cohort effects. To remedy this shortcoming, Renshaw and Haberman
\cite{renshaw2006cohort} proposed a modification of the Lee-Carter model that 
includes cohorts. It can be expressed mathematically as
\begin{equation}\label{s9e1}
\log{M}(x, t) = {a}_x + {b}_x{k}_t + {c}_x\gamma_{t-x},
\end{equation}
where the symbols have the same meaning as in Lee-Carter model. In the new
term, third on the right hand side, $\gamma_{t-x}$ captures the cohort effect.
It is a function of $t - x$, the (approximate) year of birth and ${c}_x$
is just the regression coefficient of this term. Although the model fitted the
male mortality data in England and Wales quite well it turned to not as 
robust as other models. The lack of robustness is believed to be related to
the shape of the likelihood function. It has more than one local maximum.
Therefore, when we change the age or the year range the optimizer ends up
settling on a different extremum each time\cite{cairns2008modelling}. Currie
simplified the Renshaw-Haberman model to
\begin{equation}\label{s9e2}
\log{M}(x, t) = {a}_x + {k}_t + \gamma_{t-x}.
\end{equation}
The simplification solved the problem of robustness but made the fit poorer
\cite{plat2009stochastic}. Plat introduced a model to address the deficiencies
of all the previous models while retaining their benefits 
\cite{plat2009stochastic}. Its mathematical structure is
\begin{equation}\label{s9e3}
\log M(x,t) = a_x + \kappa^1_t + \kappa^2_t(\bar{x} - x) + 
\kappa^3_t(\bar{x} - x)^+ + \gamma_{t-x},
\end{equation}
where $(\bar{x} - x)^+ = \max(\bar{x} - x, 0)$ and $\bar{x}$ is the mean age
in the sample range. The term $a_x$ plays a 
similar role as it did in Lee-Carter model. It controls the general shape
of the mortality curve. Of the four stochastic factors, $\kappa_t^1$
represents the change in the level of mortality across all ages, $\kappa_t^2$
represents the change in the level of mortality between ages, $\kappa_t^3$
captures the differences in mortality behavior in the early adulthood stage
and $\gamma_{t-x}$ captures the cohort effect. If one is interested in 
modeling only the advanced ages then one need not worry about the `accident
hump' of young adulthood and use the simplified model,
\begin{equation}\label{s9e4}
\log M(x,t) = a_x + \kappa^1_t + \kappa^2_t(\bar{x} - x) + \gamma_{t-x},
\end{equation}
Like the Lee-Carter model, this one too suffers from the `identifiability
problem'. For constants $c_1, c_2, d$, the set of parameters
\begin{eqnarray*}
\bar{\gamma}_{t-x} &=& \gamma_{t-x} + c_1 + c_2(t - x) \\
\bar{\kappa}_t^1 &=& \kappa^1_t + c_1 - d\bar{x}c_2 \\
\bar{\kappa}_t^2 &=& \kappa^2_t + dc_2 \\
\bar{a}_x &=& a_x + (1 - d)c_2
\end{eqnarray*}
also lead the the same values for $\log M(x, t)$. The problem is mitigated 
by requiring that 
\begin{eqnarray*}
\sum_{i=x_0}^{x_1} \gamma_i &=& 0 \\
\sum_{i=x_0}^{x_1} i\gamma_i &=& 0 \\
\sum_t\kappa_t^3 &=& 0,
\end{eqnarray*}
where $x_0$ and $x_1$ are the earliest and the latest years of birth to which
the cohort effect is fitted.

The Plat model is a synthesis of several stochastic mortality models aimed to
retain their advantages and eliminating their shortcomings. Its four stochastic
factors ensure that model is not trivial and yet it is not too complex. It
retains the correlation structure between age groups, making it a valuable
tool for solvency calculations. The model is applicable over the full range of 
human mortality. It captures the cohort effect and has a robust numerical 
behavior. Another valuable feature of the model is that one can develop its
risk-neutral version to price longevity derivatives.

\section{Cause-of-Death models}\label{s11}
All mortality models considered so far have considered only the occurrence of
death without examining the reason. Unlike most of the animal world, death in
humans is not solely due to natural causes. The cause-of-death models split the
aggregate mortality data into a number of causes and model mortality for each
cause separately. It as important to understand the key drivers of human 
mortality as the aggregate mortality rate. It has been observed that deaths
due to cardiovascular diseases have reduced in past few decades. Death rate
due to neoplasm have also begun to drop. However, deaths due to dementia in 
advanced ages and drug-abuse in the younger ages have risen. Features like this
do not emerge by modeling the aggregate death rate. Boumezoued and others
\cite{alexandre2019modeling} used Human Mortality Database and Human Cause-of-
death database to construct a cause-of-death mortality model for the American
population. They define $\mu_{x,t,i}$, the death rate for an age group $x$ due
to cause $i$ in year $t$ as
\begin{equation}\label{s11e1}
\mu_{x,t,i} = \frac{D_{x,t,i}}{E_{x,t}},
\end{equation}
where $D_{x,t,i}$ is the number of deaths by cause $i$ in year $t$ of an 
individual aged $x$ last birthday. $E_{x,t}$ is the total life duration in the
year $t$ of individuals ages $x$ last birthday. It is also called the 
`exposure-to-risk'. Note that the exposure is not indexed by $i$ and is an
aggregate metric. The authors use the competing risks framework according to
which every cause of death has a cause-specific lifetime associated with  it.
Thus if $\tau^A$ is the lifetime under cause $A$ and $\tau^B$ is the lifetime
under all causes except $A$ then the survival function at age $a$ is defined
as the probability that lifetimes under all clauses will be higher than $a$,
Thus,
\begin{equation}\label{s11e2}
S(a) = \pr(\tau > a) = \pr(\tau^A>a,\tau^B>a) = 
\exp\left(-\int_0^a\mu(y)dt\right),
\end{equation}
where
\begin{equation}\label{s11e3}
\tau = \min(\tau^A, \tau^B)
\end{equation}
If the lifetimes under cause $A$ is independent of the lifetime under all causes
then one can write the survival function as
\begin{equation}\label{s11e4}
S(a) = \pr(\tau^A>a)\pr(\tau^B>a)=
\exp\left(-\int_0^a\mu_A(y)dt\right)\exp\left(-\int_0^a\mu_B(y)dt\right),
\end{equation}
In equations \eqref{s11e2} and \eqref{s11e4}, $\mu$ is the aggregate mortality
rate and $\mu_A$ and $\mu_B$ are cause-specific mortality rates. There are 
several ways to build cause of death forecasts. One can associate a link
between the causes and a number of clinical factors and use the latter for
forecasting. However, doing so needs a deep expertise in several medical fields
and the model needs a large amount of medical data. Another way is to use
the multivariate Lee-Carter framework to get the join distribution of cause-
specific death rates. The model is defined by the equation
\begin{equation}\label{s11e5}
\log\mu_{x,t,i} = \alpha_{x,i} + \beta_{x,i}\kappa_{t,i},
\end{equation}
where $\alpha_{x,i}$ is the static, cause-specific age structure, $\kappa_{t,i}$
is the cause-specific mortality and $\beta_{x,i}$ is the sensitivity of the
age class $x$ to the cause $i$.

Cause-of-death models help us understand the interaction and joint-dynamics
of the causes of death. They allow us to understand the severity of death-
related pathologies. Further, if certain causes of death are prevented by
improvements in healthcare or better policies then one can take that into 
account while building aggregate mortality.

\section{Generalized linear models}\label{s12}
The ordinary least square linear regression assumes that the residuals are
normally distributed. The expected value of the response is a linear
combination of the predictors. Therefore a constant change in the predictor
results in a proportional change in the respose, a characteristic of a linear
relationship. However, often times this is not true. A generalized linear 
model expresses a certain function of the response as a linear model. That is,
instead of fitting a linear model between a response $y$ and the predictors $X$,
we git a linear model between a function $g(y)$ of the response and the
predictors. The function $g$ is called the link function. This method,
developed by Nelder and Wedderburn \cite{nelder1972generalized} proposes a 
general structure for several types of regression including the traditional
ordinary-least-squares, the logistic and the Poisson regressions.

If the probablity mass or density function of a random variable $Y$ can be
written as
\begin{equation}\label{s12e1}
f_Y(y, \theta; \phi) = 
\exp\left(\frac{y(\theta) - b(\theta)}{\alpha(\phi)} + c(y, \phi)\right)
\end{equation}
where $a, b, c$ are functions of their arguments and $\theta, \phi$ are
parameters then $Y$ is said to belong to an exponential family of distributions.The normal, binomial, Poisson, exponential and gamma distributions are all
members of the exponential family. A generalized linear model assumes that
the responses $Y_1, \ldots, Y_n$ are independent and identically distributed
with the distribution from the exponential family.

A generalized linear model consists of 1) a distribution from the exponential
family, 2) a linear predictor $\eta = X\beta$ and 3) a link function $g$
such that $\E(Y|X) = g^{-1}(\eta)$.

If we write $B = e^{k_1}$ in equation \eqref{s3e3} describing the force of
mortality of Gompertz' law then we can take logarithm of both sides to get
\begin{equation}\label{e12e2}
\log\mu(x) = k_1 + kx.
\end{equation}
The right hand side of this equation is a linear function in $x$ while the left
hand side is a logarithm. This is precisely the form of generalized linear
model with logarithmic link function and a Poisson distrubution. If the errors
in Gompertz's law are modeled in terms of a binomial distribution then the
link function is a complimentary log-log function \cite{currie2013fitting} 
defined as
\begin{equation}\label{e12e3}
f(x) = \log(-\log(1 - x)),
\end{equation}
where $0 < x < 1$. Mortality laws expressed as sigmoid functions of the form
\begin{equation}\label{e12e4}
q_x = \frac{e^{\theta_0 + \theta_1 x}}{1 + e^{\theta_0 + \theta_1 x}},
\end{equation}
can be easily transformed to the form
\begin{equation}\label{e12e5}
\frac{q_x}{1 - q_x} = e^{\theta_0 + \theta_1 x}
\end{equation}
or, taking logarithm of both sides,
\begin{equation}\label{e12e6}
\log\frac{q_x}{1 - q_x} = \theta_0 + \theta_1 x.
\end{equation}
Once again we observe that the right hand side is a linear function of the
predictor. The function on the left hand is called a logit function and it is
a legitimate link function for a generalized linear model. In fact, this link
function is used to implement the standard logistic regression in applied
statistics. It is possible to show that the Age-Period-Cohort models and the
Cairns-Blake-Dowd models can also be expressed as generalized linear models. 
The algebra involved to demonstrate this fact is a bit involved and we refer
to the paper by Currie \cite{currie2013fitting} for the details.

Generalized linear model gives the theoretical structure to several mortality
models and also suggests a practical implementation. It is important to note
that generalized linear models were not developed to build newer mortality
forecasting models. Rather, deterministic mortality models were cast as 
generalized linear models and that permitted practitioners to use a standard
algorithm to fit their data. Unlike ordinary least squares regression which
has an exact analytical solution, none exists for the generalized linear models.
They must be solved numerically using an optimizer. The optimizer is guaranteed
to find a global minimum if its objective function is convex. When it is not
it is not uncommon for the optimizer to settle on a local minimum. We mentioned
this problem in the case of the Renshaw-Haberman model in section \ref{s9}.

\section{Spline models}\label{s13}
The Lee-Carter model of equation \eqref{s6e1} and the Renshaw-Haberman model
of equation \eqref{s9e1} contain multiplicative terms of predictors. It is 
not possible to cast them as generalized linear models. However, it is possible
to express them as generalized nonlinear models with a link function. Both
these models can be expressed in terms of Poisson distribution and a logarithmic
link function. Like their linear counterparts, generalized non-linear models
do not have an exact analytical solution. Their numerical solution needs an
optimizer. However, it is not always possible to find an optimal solution.
Currie reports that \cite{currie2013fitting} among the $25$ simulations he
tried to fit these models on UK data, in six cases the model did not converge
and in five other cases the optimizer was unable to fit the model.

Splines were used to draw a smooth curve passing through a set of points. They
were initially used by draftsmen who represented the points by heavy leads and
the splines by flexible wooden strips. This idea was borrowed into numerical
analysis by I. J. Shoenberg \cite{schoenberg1969cardinal} in the late sixties
of the twentieth century. A set of polynomials were used to approximate a
non-linear function where each polynomial fitted a portion of the function
being approximated. It was possible to show that under certian conditions
\cite{schoenberg1969cardinal} a non-linear function could be expressed as a 
piecewise polynomial function. The polynomials were required to blend into each
other that the boundaries in a smooth way. That is, not only were their values
required to match but also their derivatives.

Splines can be implemented using the Cox-de Boor recurrence relation
\cite{de1978practical}
\begin{equation}\label{s13e1}
B_{i,0}(x) = \begin{cases}
1 & \text{ if } x \in [t_i, t_{i+1}] \\
0 & \text{ otherwise}
\end{cases}
\end{equation}
and
\begin{equation}\label{s13e2}
B_{i,j}(x) = \frac{x - t_i}{t_{i+j} - t_i}B_{i,j-1}(x) + 
\frac{t_{i+j+1}-x}{t_{i+j+1}-t_{i+1}}B_{i+1,j-1}(x).
\end{equation}
